---
title: "Climate Visualizations"
output: github_document
---

### Intoduction
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(maptools) #the maptools library will be used to generate the world map to plot over
data("wrld_simpl") 
library(tidyverse)

climate <- read_csv('climate.csv')
```

The original dataset is 59,223 rows and 75 columns. Missing values are represented at -9999
```{r}

climate[1:7, 1:7]
```

## Missing Values

Lets see how many missing values we have and whether they are constant across time
```{r, message=FALSE}
climate <- climate %>% 
              mutate(date = str_c(year, month, 1, sep= '-')) #this column will be useful when we attempt to plot data across time

climate[climate == -9999] <- NA 

null_df <- data.frame(is.na(climate[, 4:75])) #create boolean array to count missing values
null_df <- cbind(year=climate$year, null_df) 

null_long <- null_df %>% pivot_longer(cols = 2:73, names_to = 'column', values_to= 'isnull') #convert to long format

null_by_year <- null_long %>% group_by(year) %>% dplyr::summarise(percent_null = mean(isnull)*100) #average amount of null values per year

ggplot(data=null_by_year, aes(x=year, y=percent_null)) +geom_point() + theme_minimal() + labs(x= 'Year', y= '% Missing')
```

## Data Cleaning and Prep

It is clear from the graph above that the data has become more complete over time. The data completeness was sporadic prior to the 1950s but has become very consistent since. The outlier in the final year in which data was provided (2016) is due to it being an incomplete year at the time the data was formed. We could remove this year but since the data is already adjusted based on the location and month we will leave it as is.

Well convert the original data frame to long format so that each row is made up of a unique combination of date, latitude, and longitude values. We will remove rows with missing values and separate the latitude and longitude values into. 
```{r}
climate <- climate %>%
          pivot_longer(cols = 4:75, names_to = "lon", values_to = 'temp') %>% #convert to long format
          filter(temp != -9999) %>%
          mutate(latlon = str_c(lat, lon, sep=' ' )) %>% #this coordinate key will be useful later as we group by location
          mutate(temp= temp/100) %>% #convert temperature from hundredths of degrees Celsius to the more common, degrees Celsius
          separate(col=lon, into=c('scrap', 'lon1', 'lon2'),  remove=TRUE ) %>% 
          separate(col=lat, into=c('lat1', 'lat2'),  remove=TRUE ) %>% 
  
          select(-scrap)

head(climate)
```



## Introductory Visualizations

Here we will visualize how global temperatures have been trending since the 1880s.
```{r, message=FALSE}
monthly <- climate %>% group_by(date) %>% summarise(avg_deviance = mean(temp))

mid = mean(monthly$avg_deviance)

ggplot(data=monthly, 
       aes(x=date, y=avg_deviance, color= avg_deviance )) +
       geom_point() + 
       labs(title= 'Average Global Temperature', 
                 x='1880 - 2016', 
                 y= 'Average Global Temperature Deviance =(°C))') + 
       theme(
                axis.text.x = element_blank(),
                axis.ticks = (element_blank())) +

        scale_color_gradient2(low='blue', mid='orange', high='red', midpoint = mid)
```

## Refining Coordinates

To be able to plot our coordinates onto a world map they will need to be in numerical format, converting 
northern and eastern coordinates to positive numericals and Southern and Western coordinates to negative numericals. 
```{r}

climate <- climate %>%
          mutate(lon2 = if_else(str_sub(lon2, -1) == 'E',
                                str_sub(lon2, start=1, end=-2), #if East leave as positive and remove the 'E'
                                paste('-', (str_sub(lon2, start=1, end=-2)), sep=''))) %>% # if West convert to negative and remove the 'W'
  
          mutate(lat2 = if_else(str_sub(lat2, -1) == 'N',
                                str_sub(lat2, start=1, end=-2), #if North leave as positive and remove the 'N'
                                paste('-', (str_sub(lat2, start=1, end=-2)), sep=''))) %>% # if South convert to negative and remove the 'S'

          mutate(lon2 = as.numeric(lon2),
                 lat2 = as.numeric(lat2),
                 lat1 = if_else(lat2 >= 0, lat2 - 5, lat2+5), 
                 lon1 = if_else(lon2 >= 0, lon2 - 5, lon2+5))

```


```{r}

climate <- climate %>% pivot_longer(cols= c('lat1', 'lat2'), names_to = 'lat_number', values_to= 'lat') %>%
                       pivot_longer(cols= c('lon1', 'lon2'), names_to = 'lon_number', values_to= 'lon') 
                     # select(-scrap, -scrap2)
head(climate)
```

```{r}
#We will add a decade variable for later grouping
climate <- climate %>% 
        mutate(decade = str_c((year - (year %% 10)), 's'))
```

## Making Data Compatible with MapTools

We must order the data in the order we want the vertices to be plotted. In this case we would want them to be in a clock-wise order.
Source: https://rstudio-pubs-static.s3.amazonaws.com/86115_e78c3a8e3ec9446892a3bc1838e170c4.html
```{r}
climate <- climate %>%
  group_by(latlon) %>%
  group_by(lat_number) %>%
  arrange(ifelse(lat_number=="lat1",lon,-lon)) %>%
  ungroup() %>%
  arrange(latlon) %>%
  select(-lat_number, -lon_number) # we no longer need the latitude and longitude numbers
```

## Global Temperatures by Decade

```{r, fig.width=12, fig.height=94, message=FALSE}

ggplot() +
    theme_bw() +
    geom_polygon(data=wrld_simpl, #this generates the world map in the background
                 aes(x=long, 
                     y=lat, 
                     group=group), 
                 fill=NA, color="black") + 
    geom_polygon(data= climate,
                 aes(x=lon,
                     y=lat,
                     group=latlon,
                     fill=temp),
                 alpha=0.88, ) +  
      theme(axis.text.x = element_blank(),
            axis.ticks = (element_blank()))+
    facet_wrap(~decade,  ncol=1, ) + #this ensures we're creating a separate map for each decade
    scale_fill_gradient2(limits=c(-20, 15), # handle the color gradient
                         low='blue', mid='white', high='red', midpoint = 0) +
    labs(y="latitude", x="longitude", fill="anomaly, deg. C", title='Climate by Decade')
```

## 2016 Global Temperatures

To highlight the sharp uptick in recent temperatures, we will narrow our scope to 2016, the most recent data available. 
```{r, message= FALSE}

recent_data <- climate %>% filter(year == 2016)

ggplot() +
    theme_bw() +
    geom_polygon(data=wrld_simpl, # outline of countries
                 aes(x=long, 
                     y=lat, 
                     group=group), 
                 fill=NA, color="black") +
    geom_polygon(data= recent_data, # overlay slightly translucent temperature colors
                 aes(x=lon,
                     y=lat,
                     group=latlon,
                     fill=temp),
                 alpha=0.90, ) +       
  
    theme(axis.text.x = element_blank(),
          axis.ticks = (element_blank()))+
    scale_fill_gradient2(limits=c(-20, 15), # handle the color gradient
                         low='blue', mid='white', high='red', midpoint = 0) +
    labs(y="latitude", x="longitude", fill="anomaly, deg. C", title='2016 Temperatures')
```


## Conclusions

Now were able to get a better understanding of how global climate change has impacted different areas of the world across time. First, while temperatures have clearly been rising everywhere, the 2010s show a noticeable uptick which is even more dramatic when looking solely at 2016. This is in congruence with what we saw in the scatter-plot earlier, a sharp rise in recent years. Secondly, not every region has experienced similar changes. Parts of the U.S and Eastern Europe have actually shown colder than average temperatures in recent years This accentuates the stark contrast between weather and climate and should be understood in the context of trends on the global scale. Lastly, despite improving slightly in recent years, this dataset lacks sufficient data to make any claims about the polar regions, which are arguably the most important in terms of moderating the global climate. 